---
title: "Week 6: Causes of the Industrial Revolution" 
author: | 
    | Dr. Gabriel Geisler Mesevage 
    | gabriel.mesevage@kcl.ac.uk
    | Office Hours: Tues. 9-10am & Thurs. 4-5pm
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
format:
  revealjs:
    theme: solarized
    transition: slide
    transition-speed: fast
    slide-number: true
    footer: "British Economic History"
    css: ["helper_functions/extra.css":::
    beforeInit: "helper_functions/macros.js"
---

### Today's Plan

::: {.columns}
::: {.column width="50%"}
::: {.large}
### Part 1

+ Some quick advice on essay writing
+ Counterfactuals and causes in history


 {.column width="50%"}
::: {.large}
### Part 2

+ Allen's theory of induced innovation
+ Some critiques of Allen

::: {.columns}
::: {.column width="50%"}
::: {.large}
### Part 3

+ Mokyr's theory of 'industrial enlightenment'
+ Some criticism


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,
                      message = FALSE, fig.align='center', fig.retina=3,
                      out.width="75%")

library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           style = "markdown",
           dashed = TRUE, hyperlink = FALSE)
source("helper_functions/helper_print_bib.R")
bib <- ReadBib("pres.bib")
```

```{r setup-theme, include = FALSE}
source("helper_functions/theme_lecture.R")
```

---


class: center, middle, inverse

# Part I

---

### Some brief advice on writing essays

::: {.columns}
::: {.column width="50%"}
### Answer the question!!

+ ...by presenting and interpreting other scholars answers to the question
+ This should be easy as you can pick your own question
    - Check with me first that your question is OK

### Care about your reader

+ Spelling, Style, Structure
+ Say what you will do and then do it

::: {.column width="50%"}
### Things not to worry about

+ Don't worry about your essay "not being economics-y enough"
+ You are not obliged to incorporate primary sources
    - If it is natural go ahead
+ You can use the first person "I will..."

### Give yourself time to edit!!!

> 
:::

---
### Some brief advice on writing essays: essay structure

::: {.columns}
::: {.column width="50%"}
::: {.large}
+ 2,000 words $\approx$ 5 paragraphs (400 words/paragraph)
    + Each paragraph should make a point
    + Subtracting intro and conclusion you can only make ~3 points. Make them count!
::::::

--

 {.column width="50%"}
::: {.large}
+ How to write an introduction
  + What you will argue and what you will cover
  + *Do not* put definitions
  + *Do not* put pre-history


+ How to write a conclusion
  + Summarize argument
  + Assess strengths and weaknesses


---

### Causation: a Few Thoughts

### Causality in contemporary social science

+ Revolves around the idea of a **counterfactual**
    - Had you not taken an aspirin would you still have a headache?

::: {.columns}
::: {.column width="50%"}    
```{r causality, fig.align='center', out.height="70%"}
library(DiagrammeR)

mermaid("
graph LR
  A(Aspirin) -->|Take| B(Head Pain<br>with Aspirin)
  A -->|Don't Take| C(Head Pain<br>w/out Aspirin)
")

```

::: {.column width="50%"}
+ Difference in headache status is the causal effect
+ **Can never be observed!**
    - Attributing a cause is about defining the unobserved counterfactual
    - Causal claims flow from causal stories (models/narratives)
:::
---

### Counterfactuals and history

::: {.columns}
::: {.column width="50%"}
+ Counterfactuals are debated in history
+ Much more common in economic history/social-scientific history
    - Why?

::: {.column width="50%"}

::: {.columns}
::: {.column width="50%"}
![:scale 80%:::(figures/virtual_history.jpg)
::: {.column width="50%"}
![:scale 80%:::(figures/altered_pasts.jpg)
---

::: {.inverse}, middle, center

# Part II

---

### Allen on Why the Industrial Revolution was British

::: {.columns}
::: {.column width="50%"}

```{r allen_graph, fig.align='center', fig.retina=3, out.width="100%"}
library(DiagrammeR)

mermaid("
graph LR
  A((Commercial<br>Revolution)) --> B((Wages))
  C((Geography)) --> D((Energy<br>Prices))
  B((Wages)) --> F((Innovation))
  D((Energy<br>Prices)) --> F((Innovation))
  E((Capital<br>Prices)) --> F((Innovation))
  A --> G((Machine<br>Tools))
  G --> F
  B --> H((Education/<br>Training))
  H --> F
  A --> I((Urbanization))
  I --> D
  D --> E
style I fill:#93a1a1
style A fill:#93a1a1
style B fill:#93a1a1
style C fill:#93a1a1
style D fill:#93a1a1
style E fill:#93a1a1
style F fill:#93a1a1
style G fill:#93a1a1
style H fill:#93a1a1
")


```

::: {.column width="50%"}
+ Allen's theory is multi-causal
    - High relative wages are one important component
+ He mainly wants to explain the adoption of technology in production -- industrial innovation

<br>

+ Methodologically he insists on a **comparative** view: why Britain as opposed to elsewhere?
:::

---

### Allen: Core causal mechanism

::: {.columns}
::: {.column width="50%"}
```{r allen_graph_3, fig.align='center', fig.retina=3, out.width="100%"}

mermaid("
graph LR
  A((Commercial<br>Revolution)) --> B((Wages))
  C((Geography)) --> D((Energy<br>Prices))
  B((Wages)) --> F((Innovation))
  D((Energy<br>Prices)) --> F((Innovation))
  E((Capital<br>Prices)) --> F((Innovation))
  A --> G((Machine<br>Tools))
  G --> F
  B --> H((Education/<br>Training))
  H --> F
  A --> I((Urbanization))
  I --> D
  D --> E
style I fill:#93a1a1
style B fill:#dc322f
style D fill:#dc322f
style E fill:#dc322f
style A fill:#93a1a1
style C fill:#93a1a1
style F fill:#93a1a1
style G fill:#93a1a1
style H fill:#93a1a1
")

```

::: {.column width="50%"}
+ The core elements of the theory are
    * The price of labor
    * The price of capital
    * The price of energy
    * In particular the focus is on **relative prices**
        + High wage costs and low energy/capital costs incentivize switching machines for labor
        + These are factors that effect the **demand** for innovation by firms
:::

---

### Allen: additional supply factors

::: {.columns}
::: {.column width="50%"}
```{r allen_graph_core, fig.align='center', fig.retina=3, out.width="100%"}

mermaid("
graph LR
  A((Commercial<br>Revolution)) --> B((Wages))
  C((Geography)) --> D((Energy<br>Prices))
  B((Wages)) --> F((Innovation))
  D((Energy<br>Prices)) --> F((Innovation))
  E((Capital<br>Prices)) --> F((Innovation))
  A --> G((Machine<br>Tools))
  G --> F
  B --> H((Education/<br>Training))
  H --> F
  A --> I((Urbanization))
  I --> D
  D --> E
style I fill:#93a1a1
style B fill:#93a1a1
style D fill:#93a1a1
style E fill:#93a1a1
style A fill:#93a1a1
style C fill:#93a1a1
style F fill:#93a1a1
style G fill:#dc322f
style H fill:#dc322f
")

```

::: {.column width="50%"}
+ The **supply** of innovation increased by
    + Availability of education
        - Product of high wages (as education is costly)
    + Availability of industries (like watches) linked to scientific thought and machine tools
    + For Allen these are secondary factors
+ **Warning**: If quality of labor is too good implies the **real** wage may not be high
:::

---

### Allen: causes of the causes, why were wages high?

::: {.columns}
::: {.column width="50%"}
```{r allen_graph_4, fig.align='center', fig.retina=3, out.width="100%"}

mermaid("
graph LR
  A((Commercial<br>Revolution)) --> B((Wages))
  C((Geography)) --> D((Energy<br>Prices))
  B((Wages)) --> F((Innovation))
  D((Energy<br>Prices)) --> F((Innovation))
  E((Capital<br>Prices)) --> F((Innovation))
  A --> G((Machine<br>Tools))
  G --> F
  B --> H((Education/<br>Training))
  H --> F
  A --> I((Urbanization))
  I --> D
  D --> E
style I fill:#93a1a1
style B fill:#93a1a1
style D fill:#93a1a1
style E fill:#93a1a1
style A fill:#dc322f
style C fill:#dc322f
style F fill:#93a1a1
style G fill:#93a1a1
style H fill:#93a1a1
")

```

::: {.column width="50%"}
- Thinks wages grew through external trade/colonialism
- Thinks low energy prices stem from geography: availability of coal deposits
    + Also trade drives commercial adoption of coal through urbanization
    
    <br>
    
- Low energy prices subsidize cost of capital goods (coal used in producing iron, etc.)
:::

---

### Critiques of Allen

### Jane Humphries

+ Family structure as depicted is unrealistic
    - Need more than 3x a man's subsistence wage to support wife/children
    - Therefore wage averaged over family is not high
    - Implies probably women/children working $\rightarrow$ low wage labor

--

### Judy Stephenson

+ Re-examines key wage series Allen relies on: builder's wages in London

> "...the bills that have been interpreted in the past as reports of wages received by labourers and craftsmen did not, in fact, state the pay of workers, but the rates charged by major building contractors to clients for types of service. ...Once contractors’ margins are taken into account, the actual wage in London construction was significantly below the levels reported in the series that have been used by Allen and others."

---
# Part III

---

::: {.columns}
::: {.column width="50%"}

### Mokyr, on the Intellectual Origins of the Industrial Revolution

```{r mokyr_0, fig.align='center', fig.retina=3, out.width="200%"}

mermaid("
graph LR
  A((Enlightenment)) --> B((Scientific<br>Culture))
  C((Cultures of<br>Improvement)) --> D((Industrial<br>Enlightenment))
  D --> G((Knowledge<br>Production))
  A --> C
  B --> D
  D --> E((Access<br>Costs))
  E --> F((Innovation))
  G --> F
style A fill:#93a1a1
style B fill:#93a1a1
style C fill:#93a1a1
style D fill:#93a1a1
style E fill:#93a1a1
style F fill:#93a1a1
style G fill:#93a1a1
")

```

::: {.column width="50%"}
+ **cultural** account grounded in Enlightenment
    - Cultures that produce **useful knowledge**
+ Prime movers are small fraction of high human capital individuals
+ Causal forces are broadly European
    - Britain has some advantages, but Europe would industrialize anyway
:::
---

::: {.columns}
::: {.column width="50%"}

### Mokyr, on the Intellectual Origins of the Industrial Revolution

```{r mokyr_1, fig.align='center', fig.retina=3, out.width="200%"}

mermaid("
graph LR
  A((Enlightenment)) --> B((Scientific<br>Culture))
  C((Cultures of<br>Improvement)) --> D((Industrial<br>Enlightenment))
  D --> G((Knowledge<br>Production))
  A --> C
  B --> D
  D --> E((Access<br>Costs))
  E --> F((Innovation))
  G --> F
style A fill:#93a1a1
style B fill:#93a1a1
style C fill:#93a1a1
style D fill:#dc322f
style E fill:#93a1a1
style F fill:#93a1a1
style G fill:#93a1a1
")

```

::: {.column width="50%"}
+ Core idea is a particular form of Enlightenment: **Industrial Enlightenment**
    - More focussed on science/engineering than elsewhere
    - Devoted to creating *useful knowledge*
    - Entails strong links between scientists and manufacturers
:::

---

::: {.columns}
::: {.column width="50%"}

### Mokyr, on the Intellectual Origins of the Industrial Revolution

```{r mokyr_2, fig.align='center', fig.retina=3, out.width="200%"}

mermaid("
graph LR
  A((Enlightenment)) --> B((Scientific<br>Culture))
  C((Cultures of<br>Improvement)) --> D((Industrial<br>Enlightenment))
  D --> G((Knowledge<br>Production))
  A --> C
  B --> D
  D --> E((Access<br>Costs))
  E --> F((Innovation))
  G --> F
style A fill:#93a1a1
style B fill:#93a1a1
style C fill:#93a1a1
style D fill:#93a1a1
style E fill:#dc322f
style F fill:#93a1a1
style G fill:#dc322f
")

```

::: {.column width="50%"}
+ *Knowledge production*: both scientific but also descriptive
    - Effect is more than cumulative as knowledge interacts
+ *Access costs* to information
    - Learned societies
    - Books/journals
    - Encyclopedias
    - Universities/Meeting places
:::

---

::: {.columns}
::: {.column width="50%"}

### Mokyr, on the Intellectual Origins of the Industrial Revolution

```{r mokyr_3, fig.align='center', fig.retina=3, out.width="200%"}

mermaid("
graph LR
  A((Enlightenment)) --> B((Scientific<br>Culture))
  C((Cultures of<br>Improvement)) --> D((Industrial<br>Enlightenment))
  D --> G((Knowledge<br>Production))
  A --> C
  B --> D
  D --> E((Access<br>Costs))
  E --> F((Innovation))
  G --> F
style A fill:#93a1a1
style B fill:#dc322f
style C fill:#dc322f
style D fill:#93a1a1
style E fill:#93a1a1
style F fill:#93a1a1
style G fill:#93a1a1
")

```

::: {.column width="50%"}
+ *Belief in progress* as attainable in many domains (politics, ethics, natural philosophy) seen as important cultural view
    - Supports will to experiment in production
+ *Baconian science*:
    - More empirical & concrete than e.g. Descartes
    - More closely allied to applied scientific work
:::

---

### Critiques of Mokyr

### O'Brien

+ Mokyr inexplicably dismissive of the litany of other causal factors (geography, agriculture, institutions, etc)
+ Does not engage in *reciprocal comparison* -- comparing Britain to comparable units elsewhere in the world to ensure these mechanisms are absent
+ Does not really examine influence of Enlightenment ideas elsewhere

> "I conclude that an enlightened economy... are labels awaiting research and development to mature into operational concepts."


---


### Discussion

.center[
> What, in O'Brien's view, is missing from both Mokyr and Allen's accounts?

:::

--
<br>
> Does Mokyr's approach convince us that the industrial revolution would not have occurred without the Industrial Enlightenment?

--

<br>
> Does Humphries depiction of regular hunger and immiseration matter for Allen's argument? Is her criticism convincing?

--

<br>
> What about Stephenson's revision of wage rates? Does it fatally undermine the argument?

---
::: {.columns}
::: {.column width="50%"}
### How do we interpret the Ealing gardener's annual budget?

+ Is this a lot/little?
+ What kind of questions should we ask about this?

::: {.column width="50%"}
![:scale 100%:::(figures/ealing_budget.png)
:::